---
title: "Testing of Homogeneity Of Covariance Matrix"
author: "Tanweer Shapla"
date: "`r Sys.Date()`"
output: word_document
---


```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = T, results=T, message = F, warning = F, fig.show = "nohide", comment="")
```

In checking homogeneity of covariance matrices for $K$ populations, the null and alternative hypotheses are respectively, 


$H_0: \Sigma_1=\Sigma_2=....=\Sigma_K$

vs $H_a$: at least one equality does not hold.


## Example 1

In this example, we work on Psychological test data. We use boxM() function from heplots package, which depends on car and carData packages. 

```{r}
setwd("D:/LianZuo/Applied Statistics Course Materials/STAT 577 - Applied Multivariate Statistics/data-577")
PT=read.table("Psychological tests.DAT", header=F)
colnames(PT)=c("gender", "t1","t2", "t3", "t4")
gender=as.factor(PT$gender)

#install.packages("heplots")
library(car)
library(carData)
library(heplots)
boxM(PT[, 2:5],PT[, 1])
#or, boxM(PT[, c("t1","t2", "t3", "t4")], PT[, "gender"])
```
The test statistic value is found to be 13.551 with a p-value of 0.1945. Therefore, we fail to reject the null hypothesis and conclude that the population covariance matrices are not significnatly different. 


## Example 2

We work on iris data to check the homogeneity of covariance matrices.

```{r}
names(iris)

result <- boxM(iris[, 1:4], iris[, "Species"])
result
```


